The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-09-22 adaptation/article/article.tex
Second draft

\abstract{Abstract}
 Humans are able to adapt their movements to changes in the dynamics of the environment and of their own bodies, a phenomenon known as motor adaptation. Motor adaptation has been shown to be contextual in experiments in which participants must learn to adapt to different altered dynamics, and must be able to switch between the learned adaptations dynamically. In this work, we focus on the behavioral phenomena effected by imperfect contextual information and present a hierarchical computational model for motor adaptation based on exact Bayesian inference, which explicitly takes into account contextual information and how the dynamics of context inference affect learning and action selection. We show how our model provides a unifying explanation for the effects of sensory cues and proprioceptive information on switching between tasks, the effects of environmental volatility and predictability on learning, and the effects of previously-learned adaptations on switching between tasks, in addition to the well studied savings, de-adaptation and spontaneous recovery.

\section{Introduction}
Humans are able to adapt movements to changing dynamics of the environment and of the body itself. While we must adapt to slow changes in our body dynamics, e.g. because of muscle exhaustion throughout the day, we can also adapt to sudden changes in the environment. For example, when picking up a heavy object, we must adapt our motor commands to be able to move the object around. Even before picking it up, the size and material of an object provide us with contextual information which we can use to prepare the correct posture for moving the object. However, contextual information cannot always be relied on, e.g. a big object of unknown material.

It has been shown that humans can learn multiple tasks during the course of a single experiment, as well as learn to adapt to often-opposing modifications to the dynamics of movements \citep[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Functional_1997}, although interference can occur between tasks \citep[e.g.][]{Brashers-Krug_Consolidation_1996,Sing_Reduction_2010}, which slows down learning. Additionally, humans have been shown to dynamically switch between different learned tasks \citep[e.g.][]{Davidson_Scaling_2004,Ethier_Spontaneous_2008,Lee_Dual_2009}. The mechanisms behind these switches, the specifics of how these switches manifest in behavior, and the way in which switching can be impeded or improved, are still an active topic of research.

In order to study motor adaptation, motor error is defined in experiments as the deviation between the observed movements of a participant and a baseline movement. For example, in experiments with a mechanical arm \citep[e.g.][]{Gandolfo_Motor_1996}, a correlation coefficient is defined between the movements when the mechanical arm exerts a force and the movements when it does not; the higher the correlation coefficient, the lower the error. Other examples include measuring the force profile in error-clamp trials \citep{Smith_Interacting_2006}, endpoint errors in visuomotor rotations \citep{Kim_Neural_2015} and saccade adaptation \citep{Catz_Cerebellardependent_2008}.

@@

Bayesian accounts of motor adaptation have also been previously presented, providing an alternative explanation for savings and quick de-adaptation in the form of switching between internal models of adaptation via predictions generated by these internal models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019}. These models replace the idea of explicit multiple time scales with an estimation of their own uncertainty: as the model practices the same task, the uncertainty on its estimate of the underlying modification is reduced, which lowers variability in responses, as well as increases the stability of the estimate.

While these general models of adaptation explain the most common phenomena observed in experiments, a number of known phenomena remain outside of their scope. For example, it is known that adaptation rate is reduced in situations where the environment is unstable and unpredictable \citep{Herzfeld_memory_2014}, or situations in which errors are small \citep{Marko_Sensitivity_2012} and slowly introduced \citep{Huang_Persistence_2009}. Learning new adaptations has also been found to depend on the history of adaptations learned \citep{Vaswani_Decay_2013}.

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. In mechanical arm experiments, we can think of one context in which the arm exerts no force and one in which it does. By identifying the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly identifiable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference takes the form of an evidence-accumulating process. In this work, we propose that this process can be observed in behavioral data and can be used to explain the phenomena outlined above.

@@

Central to the model is the concept of context. In our model, contexts are defined in terms of the task to be performed (see \fref{fig:model}A), the variables of the environment that are relevant to perform the task, the forward models used by the decision-making agent to perform the task, and the update mechanisms necessary to adapt these forward models to the variables of the environment. Together, these elements allow the agent to make predictions on future observations when this context is active, and these predictions are used to identify the context. For example, when lifting an object of unknown weight, an agent might have learned one context for heavy objects and one for light objects. When observing an object to be lifted, the agent can use its size and texture to estimate the weight of the object, which in turn allows the agent to infer the appropriate context and, with it, decide how to lift the object. These contextual cues (size and texture) do not necessarily identify the context (e.g. it could be a movie prop that looks like a rock).

@@

We now briefly describe the three components of the model separately. For clarity, we start with the second component (i.e. motor adaptation), in order to introduce the terminology that we use throughout this work. For an overview, \fref{fig:model}B and C shows how these three components affect each other.

@@

It is widely accepted that in order to adapt motor commands to a novel environment, we create and update internal forward models of the outcomes of control signals \citep{Wolpert_Multiple_1998}. Here, we use this principle and chose to make use of an exact Bayesian learner. As has been discussed before \cite{Oh_Minimizing_2019}, a Bayesian learner has the advantage of not only fitting experimental data on motor adaptation, but also eliminates the need for explicit multiple time scales of learning as the learner can deal with adaptation after any number of trials. As we will show below this feature will be useful to explain specific key phenomena parsimoniously. We show that this adaptation mechanism can deal with adaptation after any number of trials, ranging from a handful of trials in an experiment to the level of expertise a professional dancer has accrued through years of practicing the same movements.

To describe this component of the model, we make use of the example of reaching movements. It is now widely assumed that humans learn associations between motor commands and outcomes in different environments \cite{Wolpert_Multiple_1998}. These associations, called forward models, are used by humans for action selection. We can write this forward model as follows:
\begin{equation}
p(\vartheta | s, c) = f(s, c; \phi) \label{eqn:forward-model}
\end{equation}
where $\vartheta$ are the outcomes of an action (e.g. the final position of the hand in a reaching movement), $s$ is the current state (the current position of the hand), $c$ is a motor command and $\phi$ are the parameters of the forward model $f(\cdot)$. We assume that during learning the parameters $\phi$ have been fine tuned to produce an accurate forward model $f$ to guide our movements and it is these parameters which are updated during motor adaptation.

@@

Additionally, we use the term "learned adaptation" and similar to refer to motor outputs (e.g. reaching movements) that have already been adapted by the participant to a change in the dynamics of the environment. For example, the learned adaptation during a task with a curl force refers to the movements of the participant that already counteract the force exerted by the mechanical arm.

@@

In this section, we discuss savings, and the related phenomenon of de-adaptation, in terms of switching between forward models. This has been modeled before \citep[e.g.][]{Wolpert_Multiple_1998,Oh_Minimizing_2019} but we show that expanding forward model switching into context inference explains facets of savings that these previous models do not.

To show this, we examined multiple experimental studies in which savings is displayed. We categorized these studies based on the amount of contextual information made available to participants. In some experiments \citep[e.g.][]{Kim_Neural_2015,Lee_Dual_2009}, the context is clearly revealed to the participant. We call these cued-context experiments. In other experiments, partial information is available to participants \citep[e.g.][]{Davidson_Scaling_2004,Zarahn_Explaining_2008} in the form of large prediction errors, partial contextual information or reward prediction errors; we refer to these as partially-cued experiments.

@@

Note that, because context inference integrates information from many sources, many experiments in which no intentional, overt contextual cues are available indeed contain contextual information that the participant can use to infer the context. For example, in curl-force-field experiments with mechanical arms, the force feedback provided by the mechanical arm always gives participants a sense for the current context, although error-clamp trials are difficult to distinguish from adaptation trials based on force feedback alone. In fact, the mere act of holding the handle informs participants that something is different from baseline. Proprioceptive signals can also provide contextual information \cite{Dizio_Motor_1995,Shadmehr_Adaptive_1994}. The sudden appearance of motor errors can itself be a cue for contextual change, as long as the motor error is perceivably higher than the expected trial-by-trial variation in performance, and even an unusually long pause between two trials could indicate to participants that a new block with a potentially different context could have started \cite[e.g.][]{Ethier_Spontaneous_2008}.

To show how adaptation changes depending on the amount of contextual information, we selected three representative experiments from two studies \cite{Kim_Neural_2015,Oh_Minimizing_2019} which differ in the amount of contextual information. \cite{Kim_Neural_2015} performed a visuomotor rotation experiment with three contexts with different rotation: no rotation, 40 degrees and -40 degrees. Participants performed shooting movements in blocks of trials with the same rotation. Importantly, the identity of the current context was displayed as a visual cue (colored light), making this a cued-context experiment. Consistent with our model, the authors found that switching was immediate and accurate. In \fref{fig:oh-2019}A we show the results of simulations with our model using the parameters of the task, as well as the experimental results from \cite{Kim_Neural_2015}. The correspondence between simulations and experimental results can be seen in the switches between contexts, i.e. when the solid black line (representing the real context) switches between 40, 0 and -40; in both panels (left and right), the blue line, representing the agent's adaptation, quickly follows the switches. However, adaptation is incomplete (i.e. the blue line never reaches 40 or -40) in both experimental data and simulations until late in the experiment (around trial 500).

\begin{figure}
\centering
\includegraphics{./figures/figure_2.png}
\caption{Savings and de-adaptation. Data from our simulations (left column) compared to data from \cite{Kim_Neural_2015} and \cite{Oh_Minimizing_2019} (right column). Experimental data was, in all three experiments, averaged across all participants; in our simulations, a single simulation is shown. In all panels, the blue line represents the adaptation displayed by the agent as a function of trial number. The black lines represent the optimal adaptation, i.e. the size of the visuomotor rotation during the task. (A) Experiment by \cite{Kim_Neural_2015} with two visuomotor rotations (-40 and 40 degrees), in addition to baseline, with all possible transitions between A, (-A) and O. Participants must adapt to the rotations during shooting movements. The y-axis is in degrees; the black line represents the real visuomotor rotation, at either zero or 20 degrees, and the blue line shows the participants' adaptation (averaged across all participants), which can be understood as the rotation of the participants' shooting movements to compensate for the task's rotation. When black and blue lines match, the participant perfectly counteracted the visuomotor rotation and reached the target. (B) An experiment by \cite{Oh_Minimizing_2019} similar to A, but the contexts were not cued and only one context, with an adaptation of 20 degrees, is introduced. O-A and A-O transitions (i.e. adaptation and de-adaptation) can be observed. Blue and black lines are as in A. (C) Same experiment as (B) but with a 10 degree adaptation.} \label{fig:oh-2019}
\end{figure}

\cite{Oh_Minimizing_2019} performed partially-cued experiment similar to \cite{Kim_Neural_2015}. The authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in \cite{Oh_Minimizing_2019} showed immediate switching in the first experiment, which can be seen in terms of the blue line (adaptation) closely following the switches in the black line (real rotation). In the second experiment, switching happens more slowly, with adaptation (blue line) lagging behind the switches in real context (black line), and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large, and slow, noisy switches when it is low. This difference is explained by our model in terms of the size of the adaptation: as the size of the adaptation is lower (10 degrees), it becomes more difficult to distinguish errors made by incorrectly identifying the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

@@

\subsection{Slower adaptation in volatile and unpredictable environments}
One central element in our model is that adaptation is gated by the uncertainty on the current context: a prediction error observed at trial $t$ will effect motor adaptation with a magnitude proportional to the size of the error, where the proportionality constant is related to the uncertainty over the context in which the error was observed: the higher the uncertainty, the lower the size of the adaptation.

@@

\begin{figure}
\centering
\includegraphics{./figures/figure_4.png}
\caption{Motor error when switching back to a previously learned adaptation. Following the block structure from \cite{Davidson_Scaling_2004}, the first 40 trials of the switch from +8 (yellow) or -8 (blue) to 4 is shown. (A) Simulations from our model, with the task parameters from \cite{Davidson_Scaling_2004}, running eight simulations per group, each simulation representing one participant. On the y axis, the error is shown in arbitrary units, related to centimeters through a monotonically increasing transformation with the same origin. The shaded area is the SEM across simulated runs (participants). (B) Data adapted from \cite{Davidson_Scaling_2004}, showing the error in the same trials as (A).}
\label{fig:davidson-2004}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./figures/figure_5.png}
\caption{Adaptation during error clamp trials. (A) Simulated adaptation during the error-clamp trials for the three groups of participants in \cite{Vaswani_Decay_2013}, using the same colors. The groups differ in the sequence of adaptations: 1.1 performed an A-E sequence; 1.2, O-A-E; 1.3, (-A/2)-A-E; and 1.4, (-A)-E. Following \cite{Vaswani_Decay_2013}, group 1.4 is not shown in A and B, as their behavior is identical to group 1.1. The solid line is the average across 10 runs (i.e. a group of 10 simulated participants) and the shaded area represents the standard deviation. The vertical dashed line is the start of the error-clamp trials. (B) Corresponding experimental data adapted from \cite{Vaswani_Decay_2013}, figure 2. (C) Simulations: Inference over the current context, where contexts are color coded: black for baseline, orange for the counter-clockwise force, purple for the clockwise force and brown for clockwise force with half strength. The lines represent the posterior probability of each context in every trial, while the background color represents the true context. An olive-colored background represents error-clamp trials. As in (A), solid lines represent the average across all runs and shaded areas represent the standard deviation. (D) Simulations: Visualization of the lag before a change in context is detected by the agent during the E trials. Each line represents one run (groups 1.1, 1.2 and 1.3, 10 runs per group).}
\label{fig:vaswani-2013}
\end{figure}

This effect is explained in our model in terms of context inference. In \fref{fig:vaswani-2013}C, the inference over context is shown for each group separately. Context inference works reliably until the error-clamp trials start, which do not correspond to any of the known contexts. This causes the agent to infer the combination of some of the known contexts that best fits the observations. Depending on the contexts previously learned by the agent (which change from group to group), inference during E trials will vary: groups 1.1 and 1.4 have exactly the same behavior, where the previous context (A and -A, respectively) slowly dwindles. These agents will slowly lower the force applied until eventually reaching zero (after hundreds of trials; not shown). In contrast, group 1.3 has learned the additional -A/2 context (represented by brown in \fref{fig:vaswani-2013}C), which has a non-zero posterior probility during E trials, pushing the agent's adaptation force more quickly towards zero. Group 1.2 behaves similarly to 1.1, with the exception that the baseline context, which was recently seen, plays a bigger role during E trials, making the agent reduce its force during E trials slightly more slowly than groups 1.1 and 1.4.

Our model also reproduces the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context (A for groups 1.1, 1.2 and 1.3, -A for 1.4) for each simulated run (one line per run, 10 runs per group, all groups). It can be seen that the moment in which the drop starts, as well as the speed of the drops, depends on the run, reproducing the different lags observed in experimental data. This can also be seen directly in context inference, displayed as the shaded areas in \fref{fig:vaswani-2013}C.

@@

Similar models have been presented before \cite[e.g.][]{Wolpert_Multiple_1998,Baddeley_System_2003}, which infer the internal models (forward and backward) that best explain the observations and use them to issue motor command. We built on these studies in two important ways: (1) we generalized the inference over the best-fitting internal models to the concept of context inference, which includes all sensory information, in addition to feedback provided by the experimenters or the task; (2) we showed how this seemingly small generalization provides a parsimonious explanation for many experimental phenomena across many experiments in motor adaptation that, up to now, had required multiple \textit{ad hoc} mechanisms.

@@

In other cases, however, the context might not be directly observable, and context inference takes the form of an evidence-accumulating process that can take any amount of time to be sure of the context. It is in these cases where the effects of context inference are most noticeable \todo{Maybe move all text until here to the introduction}. While many experiments exist that give probabilistic contextual information \cite[e.g.][]{Scholz_uncontrolled_1999,Behrens_Learning_2007,Nassar_Dissociable_2019}, evidence accumulation is not limited to these explicit cases. Indeed, as we argued in the Results section, many experiments inadvertently include partial contextual information used by participants.

@@

The same rationale explains the results from \cite{Davidson_Scaling_2004}: motor learning, which in our model is gated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1).

>>>>

\abstract{Abstract}
 Humans are able to adapt their movements to changes in the dynamics of the environment and of their own bodies, a phenomenon known as motor adaptation. Motor adaptation has been shown to be contextual in experiments in which participants must learn to adapt to different modified dynamics, and must be able to switch between the learned adaptations dynamically. In real life, the information that we use to determine how to perform a task is often incomplete or misleading, which affects our ability to switch between tasks. In this work, we focus on the behavioral phenomena effected by imperfect contextual information and present a hierarchical computational model for motor adaptation based on exact Bayesian inference, which explicitly takes into account contextual information and how the dynamics of context inference affect learning and action selection. We show how our model provides a unifying explanation for the effects of sensory cues and proprioceptive information on switching between tasks, the effects of environmental volatility and predictability on learning, and the effects of previously-learned adaptations on switching between tasks, in addition to the well studied savings, de-adaptation and spontaneous recovery.

\section{Introduction}
Humans are able to adapt movements to changing dynamics of the environment and of the body itself. We adapt to slow changes in our body dynamics, e.g. because of muscle exhaustion throughout the day, but we can also adapt to sudden changes in the environment. For example, when picking up a heavy object, we must adapt our motor commands to be able to move the extra weight safely. Even before picking it up, the size and material of an object provide us with contextual information which we can use to prepare the correct posture for moving the object. However, contextual information cannot always be relied on, e.g. a big object of unknown material.

It has been shown that humans can learn multiple tasks during the course of a single experiment, as well as learn to adapt to often-opposing modifications to the dynamics of movements \citep[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Functional_1997}, although interference can occur between tasks \citep[e.g.][]{Brashers-Krug_Consolidation_1996,Sing_Reduction_2010}, which slows down learning. Additionally, humans have been shown to dynamically switch between different learned tasks \citep[e.g.][]{Davidson_Scaling_2004,Ethier_Spontaneous_2008,Lee_Dual_2009}. The mechanisms behind these switches, the specifics of how these switches manifest in behavior, and the ways in which switching can be impeded or improved, are still an active topic of research.

In order to study motor adaptation, motor error is defined in experiments as the deviation between the observed movements of a participant and a baseline movement. For example, in experiments with a mechanical arm \citep[e.g.][]{Gandolfo_Motor_1996}, a correlation coefficient can be defined between the movements when the mechanical arm exerts a force and the movements when it does not; the higher the correlation coefficient, the lower the error. Other examples include measuring the force profile in error-clamp trials \citep{Smith_Interacting_2006}, endpoint errors in visuomotor rotations \citep{Kim_Neural_2015} and saccade adaptation \citep{Catz_Cerebellardependent_2008}.

@@

Bayesian accounts of motor adaptation have also been presented, providing an alternative explanation for savings and quick de-adaptation in the form of switching between internal models of adaptation via predictions generated by these internal models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019}. These models replace the idea of explicit multiple time scales with an estimation of their own uncertainty: as the model practices the same task, the uncertainty on its estimate of the underlying modification is reduced, which lowers variability in responses and increases the stability of the estimate.

While these general models of adaptation explain the most common phenomena observed in experiments, other known phenomena remain outside of their scope. For example, it is known that adaptation rate is reduced in situations where the environment is unstable and unpredictable \citep{Herzfeld_memory_2014}, or situations in which errors are small \citep{Marko_Sensitivity_2012} and slowly introduced \citep{Huang_Persistence_2009}. Learning new adaptations has also been found to depend on the history of adaptations learned \citep{Vaswani_Decay_2013}.

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. In mechanical arm experiments, we can think of one context in which the arm exerts no force and one in which it does. By inferring the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly inferable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference takes the form of an evidence-accumulating process. In this work, we propose that this process can be observed in behavior and can be used to explain the phenomena outlined above.

@@

Central to the model is the concept of context. In our model, contexts are defined in terms of the task to be performed (see \fref{fig:model}A), the variables of the environment that are relevant to perform the task, the forward models used by the decision-making agent to perform the task, and the update mechanisms necessary to adapt these forward models to the variables of the environment. Together, these elements allow the agent to make predictions on future observations when this context is active, and these predictions are used to infer the context. For example, when lifting an object of unknown weight, an agent might have learned one context for heavy objects and one for light objects. When observing an object to be lifted, the agent can use its size and texture to estimate the weight of the object, which in turn allows the agent to infer the appropriate context and, with it, decide how to lift the object. These contextual cues (size and texture) do not necessarily identify the context (e.g. it could be a movie prop that looks like a rock) and integration over time might be necessary.

@@

We now briefly describe the three components of the model separately. For clarity, we start with the second component (i.e. motor adaptation), in order to introduce the terminology that we use throughout this work. For an overview, we show in \fref{fig:model}B and C how these three components affect each other.

@@

It is widely accepted that in order to adapt motor commands to a novel environment, we create and update internal forward models of the outcomes of control signals \citep{Wolpert_Multiple_1998}. Here, we use this principle and chose to make use of an exact Bayesian learner. As has been discussed before \cite{Oh_Minimizing_2019}, a Bayesian learner has the advantage of not only fitting experimental data on motor adaptation, but also eliminating the need for explicit multiple time scales of learning as the learner can deal with adaptation after any number of trials. As we will show below this feature will be useful to explain specific key phenomena parsimoniously. We show that this adaptation mechanism can deal with adaptation after any number of trials, ranging from a handful of trials in an experiment to the level of expertise a professional dancer has accrued through years of practicing the same movements.

To describe this component of the model, we make use of the example of reaching movements. We define a forward model as follows:
\begin{equation}
p(\vartheta | s, c) = f(s, c; \phi) \label{eqn:forward-model}
\end{equation}
where $\vartheta$ are the outcomes of an action (e.g. the final position of the hand in a reaching movement), $s$ is the current state (the current position of the hand), $c$ is a motor command and $\phi$ are the parameters of the forward model $f(\cdot)$. We assume that during learning the parameters $\phi$ have been fine tuned to produce an accurate forward model $f$ to guide our movements and it is these parameters which are updated during motor adaptation.

@@

In some experiments, adaptation blocks are repeated throughout the experiments. For example, \cite{Oh_Minimizing_2019} used an O-A-O-A-O-A-O-A sequence. To differentiate the repeating blocks, the authors named them $O_i$ and $A_i$, where $i = 1, 2, ...$. Their experiment can be described as $O_1-A_1-O_2-...$. We adopt this naming convention.

We use the term "learned adaptation" and similar to refer to motor outputs (e.g. reaching movements) that have already been adapted by the participant to a change in the dynamics of the environment. For example, the learned adaptation during a task with a curl force refers to the movements of the participant that already counteract the force exerted by the mechanical arm.

@@

In this section, we discuss savings, and the related phenomenon of de-adaptation, in terms of switching between forward models. This has been modeled before \citep[e.g.][]{Wolpert_Multiple_1998,Oh_Minimizing_2019,Imamizu_Explicit_2007} but we show that expanding forward model switching into context inference explains facets of savings that these previous models do not.

To show this, we examined multiple experimental studies in which savings are observed. We categorized these studies based on the amount of contextual information made available to participants. In some experiments \citep[e.g.][]{Kim_Neural_2015,Lee_Dual_2009}, the context is clearly revealed to the participant. We call these cued-context experiments. In other experiments, partial information is available to participants \citep[e.g.][]{Davidson_Scaling_2004,Zarahn_Explaining_2008} in the form of large prediction errors, partial contextual information or reward prediction errors; we refer to these as partially-cued experiments.

@@

Because context inference integrates information from many sources, many experiments in which no intentional, overt contextual cues are available indeed contain contextual information that the participant can use to infer the context. For example, in curl-force-field experiments with mechanical arms, the force feedback provided by the mechanical arm always gives participants a sense for the current context, although error-clamp trials are difficult to distinguish from adaptation trials based on force feedback alone. In fact, the mere act of holding the handle informs participants that something is different from baseline. Proprioceptive signals can also provide contextual information \cite{Dizio_Motor_1995,Shadmehr_Adaptive_1994}. The sudden appearance of motor errors can itself be a cue for contextual change, as long as the motor error is perceivably higher than the expected trial-by-trial variation in performance, and even an unusually long pause between two trials could indicate to participants that a new block with a potentially different context could have started \cite{Ethier_Spontaneous_2008}.

To show how adaptation changes depending on the amount of contextual information, we selected three representative experiments from two studies \cite{Kim_Neural_2015,Oh_Minimizing_2019} which differ in the amount of contextual information. \cite{Kim_Neural_2015} performed a cued-context visuomotor rotation experiment with three contexts with different rotation: no rotation, 40 degrees and -40 degrees. Participants performed shooting movements in blocks of trials with the same rotation. Importantly, the identity of the current context was displayed as a visual cue (colored light), making this a cued-context experiment. Consistent with our model, the authors found that switching was immediate and accurate. In \fref{fig:oh-2019}A we show the results of simulations with our model using the parameters of the task, as well as the experimental results from \cite{Kim_Neural_2015}. The correspondence between simulations and experimental results can be seen in the switches between contexts, i.e. when the solid black line (representing the real context) switches between 40, 0 and -40; in both panels (left and right), the blue line, representing the agent's adaptation, quickly follows the switches.

\begin{figure}
\centering
\includegraphics{./figures/figure_2.png}
\caption{Savings and de-adaptation. Data from our simulations (left column) compared to data adapted from figure 2A by \cite{Kim_Neural_2015} and figure 4A from \cite{Oh_Minimizing_2019} (right column). Experimental data was, in all three experiments, averaged across all participants; in our simulations, a single simulation is shown. In all panels, the blue line represents the adaptation displayed by the agent as a function of trial number. The black lines represent the optimal adaptation, i.e. the size of the true visuomotor rotation during the task. (A) Experiment by \cite{Kim_Neural_2015} with two visuomotor rotations (-40 and 40 degrees), in addition to baseline, with all possible transitions between A, (-A) and O. Participants must adapt to the rotations during shooting movements, and the blue line can be understood as the rotation of the participants' shooting movements to compensate for the task's rotation. When black and blue lines match, the participant perfectly counteracted the visuomotor rotation and reached the target. (B) An experiment by \cite{Oh_Minimizing_2019} similar to A, but the contexts were not cued and only one context, with an adaptation of 20 degrees, is introduced. O-A and A-O transitions (i.e. adaptation and de-adaptation) can be observed. Blue and black lines are as in A. Different blocks with the same adaptation are differentiated by the subscript (e.g. $A_1$ and $A_2$ have the same 20 degree visuomotor rotation). (C) Same experiment as (B) but with a 10 degree adaptation.} \label{fig:oh-2019}
\end{figure}

Participants did not completely adapt to the rotation during the first block, as evidenced by the blue line not reaching as low or high as the black line during the first blocks. Switching between opposite adaptations was nevertheless immediate in simulations and experimental data. After around 500 trials, participants had adapted completely, as seen by the blue line reaching the heights of the black line, and switches continued to be immediate.

These results are similar to those shown by \cite{Imamizu_Explicit_2007}, where sensory cues of varying reliability effected immediate or slow contextual switches. To expand on these results, we now turn to performance feedback and its effects on switching behavior.

\cite{Oh_Minimizing_2019} performed partially-cued experiments similar to \cite{Kim_Neural_2015}. The authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in the first experiment first learned the adaptation in $A_1$. From the $A_1 - O_2$ transition onward, participants showed immediate switching (with a one-trial lag) between A and O (both ways), which can be seen in terms of the blue line (adaptation) closely following the switches in the black line (real rotation). In the second experiment, switching happens more slowly, with adaptation (blue line) lagging behind the switches in real context (black line), and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large, and slow, noisy switches when it is low. This difference is explained by our model in terms of the size of the adaptation: as the size of the adaptation is lower (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

@@

\subsection{Slower adaptation in volatile and unpredictable environments}
One central element in our model is that adaptation is modulated by the uncertainty on the current context: a prediction error observed at trial $t$ will effect motor adaptation with a magnitude proportional to the size of the error, where the proportionality constant is related to the uncertainty over the context in which the error was observed: the higher the uncertainty, the lower the size of the adaptation.

@@

\begin{figure}
\centering
\includegraphics{./figures/figure_4.png}
\caption{Motor error when switching back to a previously learned adaptation. Following the block structure from \cite{Davidson_Scaling_2004}, the first 40 trials of the switch from +8 (yellow) or -8 (blue) to 4 is shown. (A) Simulations from our model, with the task parameters from \cite{Davidson_Scaling_2004}, running eight simulations per group, each simulation representing one participant. On the y axis, the error is shown in arbitrary units, related to centimeters through a monotonically increasing transformation with the same origin. The shaded area is the SEM across simulated runs (participants). (B) Data adapted from figure 4 by \cite{Davidson_Scaling_2004}, showing the error in the same trials as (A).}
\label{fig:davidson-2004}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./figures/figure_5.png}
\caption{Adaptation during error clamp trials. (A) Simulated adaptation during the error-clamp trials for the three groups of participants in \cite{Vaswani_Decay_2013}, using the same colors. The groups differ in the sequence of adaptations: 1.1 performed an A-E sequence; 1.2, O-A-E; 1.3, (-A/2)-A-E; and 1.4, (-A)-E. Following \cite{Vaswani_Decay_2013}, group 1.4 is not shown in A and B, as their behavior is identical to group 1.1. The solid line is the average across 10 runs (i.e. a group of 10 simulated participants) and the shaded area represents the standard deviation. The vertical dashed line is the start of the error-clamp trials. (B) Corresponding experimental data adapted from figure 2C by \cite{Vaswani_Decay_2013}. (C) Simulations: Inference over the current context, where contexts are color coded: black for baseline, orange for the counter-clockwise force, purple for the clockwise force and brown for clockwise force with half strength. The lines represent the posterior probability of each context in every trial, while the background color represents the true context. An olive-colored background represents error-clamp trials. As in (A), solid lines represent the average across all runs and shaded areas represent the standard deviation. (D) Simulations: Visualization of the lag before a change in context is detected by the agent during the E trials. Each line represents one run (10 runs per group).}
\label{fig:vaswani-2013}
\end{figure}

This effect is explained in our model in terms of context inference. In \fref{fig:vaswani-2013}C, the inference over context is shown for each group separately. Context inference works reliably until the error-clamp trials start, which do not correspond to any of the known contexts. This causes the agent to infer the combination of some of the known contexts that best fits the observations. Depending on the contexts previously learned by the agent (which change from group to group), inference during E trials will vary: groups 1.1 and 1.4 have exactly the same behavior, where the previous context (A and -A, respectively) slowly dwindles. These agents will slowly lower the force applied until eventually reaching zero (after hundreds of trials; not shown). In contrast, group 1.3 has learned the additional -A/2 context (represented by brown in \fref{fig:vaswani-2013}C), which has a non-zero posterior probability during E trials, pushing the agent's adaptation force more quickly towards zero. Group 1.2 behaves similarly to 1.1, with the exception that the baseline context, which was recently seen, plays a bigger role during E trials, making the agent reduce its force during E trials slightly more slowly than groups 1.1 and 1.4.

Our model also reproduces the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context for each simulated run (one line per run, 10 runs per group, all groups). It can be seen that the moment in which the drop starts, as well as the speed of the drops, depends on the run, reproducing the different lags observed in experimental data. This can also be seen directly in context inference, displayed as the shaded areas in \fref{fig:vaswani-2013}C.

@@

Similar models have been presented before \cite[e.g.][]{Wolpert_Multiple_1998,Baddeley_System_2003,Imamizu_Explicit_2007}, which infer the internal models (forward and backward) that best explain the observations and use them to issue motor command. We built on these studies in two important ways: (1) we generalized the inference over the best-fitting internal models to the concept of context inference, which includes all sensory information, in addition to feedback provided by the experimenters or the task; (2) we showed how this seemingly small generalization provides a parsimonious explanation for many experimental phenomena across many experiments in motor adaptation that, up to now, had required multiple \textit{ad hoc} mechanisms.

@@

In other cases, however, the context might not be directly observable, and context inference takes the form of an evidence-accumulating process that can take any amount of time to be sure of the context. It is in these cases where the effects of context inference are most noticeable. While many experiments exist that give probabilistic contextual information \cite[e.g.][]{Scholz_uncontrolled_1999,Behrens_Learning_2007,Nassar_Dissociable_2019}, evidence accumulation is not limited to these explicit cases. Indeed, as we argued in the Results section, many experiments inadvertently include partial contextual information used by participants.

@@

The same rationale explains the results from \cite{Davidson_Scaling_2004}: motor learning, which in our model is modulated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1).
